% add u''(t)=f(t), where f is a polynomial that you can exactly represent. More collocation points and match the number of basis to 8.
% next experiment is to add f(t) that is not a polynomial family
% next is laplace in 2D

% the ODE is u''+u'+u=t^2+3t+3
% initial conditions u(-1)=0, u(1)=2
% collocation points at t=-1/3, t=1/3
% the solution is u(t) = t^2 + t  (not t^2+2t: u(-1)=1-1=0 and u(1)=1+1=2)

\documentclass{article}
\usepackage{amsmath}
\usepackage{array}

\begin{document}

\section*{Collocation Method for a Second-Order ODE with Lower-Order Terms}

\subsection*{1. The Initial Problem}
We seek a solution $u(t)$ for the following boundary value problem (BVP) on the interval $t \in [-1, 1]$:
\begin{equation}
    u''(t) + u'(t) + u(t) = t^2 + 3t + 3
\end{equation}
subject to the boundary conditions:
\begin{itemize}
    \item $u(-1) = 0$
    \item $u(1) = 2$
\end{itemize}
and the interior collocation requirements:
\begin{itemize}
    \item $L[u](-1/3) = f(-1/3)$
    \item $L[u](1/3) = f(1/3)$
\end{itemize}
where $L[\,\cdot\,] := (\,\cdot\,)'' + (\,\cdot\,)' + (\,\cdot\,)$ and $f(t) = t^2 + 3t + 3$.  Evaluating
the right-hand side at the collocation points:
\begin{equation}
    f\!\left(-\tfrac{1}{3}\right) = \tfrac{1}{9} - 1 + 3 = \tfrac{19}{9}, \qquad
    f\!\left(\tfrac{1}{3}\right)  = \tfrac{1}{9} + 1 + 3 = \tfrac{37}{9}.
\end{equation}
The exact solution to this BVP is $u(t) = t^2 + t$.

\subsection*{2. Case A: Specific Power Basis}
Using the standard monomial basis $\phi = \{1, t, t^2, t^3\}$, the trial function is
$u(t) = a_0 + a_1 t + a_2 t^2 + a_3 t^3$.
We apply the operator $L[\,\cdot\,] = (\,\cdot\,)'' + (\,\cdot\,)' + (\,\cdot\,)$ to each basis
function by computing its second derivative, first derivative, and the function itself, then summing:
\begin{align*}
    L[\phi_0]
        &= \underbrace{(1)''}_0 + \underbrace{(1)'}_0 + \underbrace{1}_{1}
         = 1 \\[4pt]
    L[\phi_1]
        &= \underbrace{(t)''}_0 + \underbrace{(t)'}_{1} + \underbrace{t}_{t}
         = 1 + t \\[4pt]
    L[\phi_2]
        &= \underbrace{(t^2)''}_{2} + \underbrace{(t^2)'}_{2t} + \underbrace{t^2}_{t^2}
         = 2 + 2t + t^2 \\[4pt]
    L[\phi_3]
        &= \underbrace{(t^3)''}_{6t} + \underbrace{(t^3)'}_{3t^2} + \underbrace{t^3}_{t^3}
         = 6t + 3t^2 + t^3
\end{align*}

\paragraph{Assembling the matrix.}
The collocation method turns each condition into one row of the matrix.  The entry in
row $i$, column $j$ is obtained by asking: ``what does condition $i$ give when applied to the
$j$-th basis function $\phi_j$ alone?''  With the four conditions ordered as
(BC at $-1$), (BC at $1$), (ODE at $-\tfrac{1}{3}$), (ODE at $\tfrac{1}{3}$), the entries are:

\begin{center}
\renewcommand{\arraystretch}{1.6}
\begin{tabular}{c|cccc}
 & $\phi_0 = 1$ & $\phi_1 = t$ & $\phi_2 = t^2$ & $\phi_3 = t^3$ \\
\hline
$u(-1)$
    & $1$ & $-1$ & $1$ & $-1$ \\
$u(1)$
    & $1$ & $1$ & $1$ & $1$ \\
$L[\phi_j]\!\left(-\tfrac{1}{3}\right)$
    & $1$
    & $1 - \tfrac{1}{3} = \tfrac{2}{3}$
    & $2 - \tfrac{2}{3} + \tfrac{1}{9} = \tfrac{13}{9}$
    & $-2 + \tfrac{1}{3} - \tfrac{1}{27} = -\tfrac{46}{27}$ \\
$L[\phi_j]\!\left(\tfrac{1}{3}\right)$
    & $1$
    & $1 + \tfrac{1}{3} = \tfrac{4}{3}$
    & $2 + \tfrac{2}{3} + \tfrac{1}{9} = \tfrac{25}{9}$
    & $2 + \tfrac{1}{3} + \tfrac{1}{27} = \tfrac{64}{27}$ \\
\end{tabular}
\end{center}
Reading off each row gives the system $Ax = b$:
\begin{equation}
\begin{bmatrix}
1 & -1 & 1 & -1 \\[4pt]
1 &  1 & 1 &  1 \\[4pt]
1 & \dfrac{2}{3}  & \dfrac{13}{9}  & -\dfrac{46}{27} \\[10pt]
1 & \dfrac{4}{3}  & \dfrac{25}{9}  &  \dfrac{64}{27}
\end{bmatrix}
\begin{bmatrix} a_0 \\ a_1 \\ a_2 \\ a_3 \end{bmatrix}
=
\begin{bmatrix} 0 \\[4pt] 2 \\[4pt] \dfrac{19}{9} \\[10pt] \dfrac{37}{9} \end{bmatrix}
\end{equation}
The right-hand side entries for the boundary rows come directly from the BCs
($u(-1)=0$, $u(1)=2$); the entries for the collocation rows are the values of $f(t)$
evaluated at the respective collocation point.

The system has a unique solution:
\begin{equation}
\begin{bmatrix} a_0 \\ a_1 \\ a_2 \\ a_3 \end{bmatrix}
= \begin{bmatrix} 0 \\ 1 \\ 1 \\ 0 \end{bmatrix},
\end{equation}
giving $u(t) = t^2 + t$, confirming the exact answer.

The matrix is decomposable into a linear combination of Pauli strings.
In this case all 16 two-qubit Pauli strings are generally needed to represent the matrix exactly.

\subsection*{3. Case B: Generalized Basis with Coefficients $C$}
We define the basis functions $\phi_i(t)$ as general polynomials with coefficients $C_{ij}$:
\begin{align*}
    \phi_0(t) &= C_{00} \\
    \phi_1(t) &= C_{11}t + C_{10} \\
    \phi_2(t) &= C_{22}t^2 + C_{21}t + C_{20} \\
    \phi_3(t) &= C_{33}t^3 + C_{32}t^2 + C_{31}t + C_{30}
\end{align*}
Applying $L[\,\cdot\,] = (\,\cdot\,)'' + (\,\cdot\,)' + (\,\cdot\,)$ to each basis function by
computing each term separately and collecting by power of $t$:
\begin{align*}
    L[\phi_0]
        &= \underbrace{0}_{\phi_0''} + \underbrace{0}_{\phi_0'} + \underbrace{C_{00}}_{\phi_0}
         = C_{00} \\[6pt]
    L[\phi_1]
        &= \underbrace{0}_{\phi_1''} + \underbrace{C_{11}}_{\phi_1'} + \underbrace{C_{11}t + C_{10}}_{\phi_1}
         = (C_{10} + C_{11}) + C_{11}\,t \\[6pt]
    L[\phi_2]
        &= \underbrace{2C_{22}}_{\phi_2''} + \underbrace{2C_{22}t + C_{21}}_{\phi_2'}
           + \underbrace{C_{22}t^2 + C_{21}t + C_{20}}_{\phi_2} \\
        &= (C_{20} + C_{21} + 2C_{22}) + (C_{21} + 2C_{22})\,t + C_{22}\,t^2 \\[6pt]
    L[\phi_3]
        &= \underbrace{6C_{33}t + 2C_{32}}_{\phi_3''}
           + \underbrace{3C_{33}t^2 + 2C_{32}t + C_{31}}_{\phi_3'}
           + \underbrace{C_{33}t^3 + C_{32}t^2 + C_{31}t + C_{30}}_{\phi_3} \\
        &= (C_{30} + C_{31} + 2C_{32})
           + (C_{31} + 2C_{32} + 6C_{33})\,t
           + (C_{32} + 3C_{33})\,t^2
           + C_{33}\,t^3
\end{align*}

\paragraph{Assembling the matrix.}
Each matrix entry $(i,j)$ is obtained by evaluating the condition for row~$i$ on the
basis function $\phi_j$ alone.  For the two boundary rows this means evaluating $\phi_j$
at the boundary point; for the two collocation rows it means evaluating $L[\phi_j]$ at
the collocation point.

\smallskip
\noindent\textit{Boundary rows.}  Since $\phi_j(t)$ is a polynomial, evaluation at
$t=-1$ and $t=1$ just substitutes the sign:
\begin{align*}
    \phi_j(-1) &: \text{ replace } t \to -1 \text{ in } \phi_j \\
    \phi_j(1)  &: \text{ replace } t \to  1 \text{ in } \phi_j
\end{align*}

\noindent\textit{Collocation rows.}  Evaluate the polynomial $L[\phi_j](t)$ derived above
at $t = -1/3$ and $t = 1/3$, collecting the resulting scalar.  For example, column $j=1$:
\begin{align*}
    L[\phi_1]\!\left(-\tfrac{1}{3}\right)
        &= \left(C_{10} + C_{11}\right) + C_{11}\!\left(-\tfrac{1}{3}\right)
         = C_{10} + C_{11} - \tfrac{C_{11}}{3}
         = C_{10} + \tfrac{2}{3}C_{11} \\
    L[\phi_1]\!\left(\tfrac{1}{3}\right)
        &= \left(C_{10} + C_{11}\right) + C_{11}\!\left(\tfrac{1}{3}\right)
         = C_{10} + C_{11} + \tfrac{C_{11}}{3}
         = C_{10} + \tfrac{4}{3}C_{11}
\end{align*}
The same procedure applied to $\phi_2$ and $\phi_3$ produces the remaining entries.
Applying the four conditions to the trial function $u(t) = \sum_{i=0}^3 a_i \phi_i(t)$ yields the system:
\begin{equation}
\begin{bmatrix}
C_{00}
  & C_{10} - C_{11}
  & C_{20} - C_{21} + C_{22}
  & C_{30} - C_{31} + C_{32} - C_{33} \\[4pt]
C_{00}
  & C_{10} + C_{11}
  & C_{20} + C_{21} + C_{22}
  & C_{30} + C_{31} + C_{32} + C_{33} \\[4pt]
C_{00}
  & C_{10} + \tfrac{2}{3}C_{11}
  & C_{20} + \tfrac{2}{3}C_{21} + \tfrac{13}{9}C_{22}
  & C_{30} + \tfrac{2}{3}C_{31} + \tfrac{13}{9}C_{32} - \tfrac{46}{27}C_{33} \\[6pt]
C_{00}
  & C_{10} + \tfrac{4}{3}C_{11}
  & C_{20} + \tfrac{4}{3}C_{21} + \tfrac{25}{9}C_{22}
  & C_{30} + \tfrac{4}{3}C_{31} + \tfrac{25}{9}C_{32} + \tfrac{64}{27}C_{33}
\end{bmatrix}
\begin{bmatrix} a_0 \\ a_1 \\ a_2 \\ a_3 \end{bmatrix}
=
\begin{bmatrix} 0 \\ 2 \\ \tfrac{19}{9} \\ \tfrac{37}{9} \end{bmatrix}
\end{equation}

\subsection*{4. Why a Block-Diagonal Structure is No Longer Available}

In Experiment~1, the operator $L_1[\,\cdot\,] = (\,\cdot\,)''$ annihilated every polynomial of degree
$\leq 1$, so the collocation rows (rows 3--4) automatically had zeros in the first two columns for
\emph{any} choice of $\phi_0$ and $\phi_1$.  This gave a lower-left zero block by construction,
and the Pauli-minimisation strategy then zeroed the upper-right block by constraint, producing a
block-diagonal matrix with at most 8 non-zero Pauli strings.

In Experiment~2, the operator $L_2[\,\cdot\,] = (\,\cdot\,)'' + (\,\cdot\,)' + (\,\cdot\,)$ contains
the identity term $(\,\cdot\,)$, so $L_2[\phi_0] = C_{00}$ and $L_2[\phi_1]$ involves both constant
and $t$-terms.  Consequently, the lower-left $2\times 2$ block of the Case~B matrix has entries
\begin{equation}
    \begin{pmatrix}
        C_{00} & C_{10} + \tfrac{2}{3}C_{11} \\
        C_{00} & C_{10} + \tfrac{4}{3}C_{11}
    \end{pmatrix}.
\end{equation}
Forcing this block to zero requires $C_{00} = 0$, $C_{10} + \tfrac{2}{3}C_{11} = 0$, and
$C_{10} + \tfrac{4}{3}C_{11} = 0$; subtracting the last two equations gives $C_{11} = 0$, then
$C_{10} = 0$, and finally $C_{00} = 0$.  But this makes $\phi_0 \equiv 0$ and $\phi_1 \equiv 0$,
so the basis is degenerate.

\paragraph{Consequence.}
The lower-left zero block that underpinned the Experiment~1 strategy is structurally
impossible in Experiment~2: any non-degenerate basis leaves the collocation rows coupled to
$\phi_0$ and $\phi_1$.  Minimising the Pauli count therefore requires a different approach
(e.g.\ an alternative $2+2$ row partition, a $1+3$ or $3+1$ partition, or operator
pre-conditioning) and may not reach three Pauli strings.

\end{document}
