\documentclass{article}
\usepackage{amsmath}
\usepackage{array}

\begin{document}

\section*{Collocation Method for a Second-Order ODE}

\subsection*{1. The Initial Problem}
We seek a solution $u(t)$ for the following boundary value problem (BVP) on the interval $t \in [0, 1]$:
\begin{equation}
    u''(t) = 0
\end{equation}
subject to the boundary conditions:
\begin{itemize}
    \item $u(0) = 0$
    \item $u(1) = 1$
\end{itemize}
and the interior collocation requirements:
\begin{itemize}
    \item $u''(1/3) = 0$
    \item $u''(2/3) = 0$
\end{itemize}

\subsection*{2. Case A: Specific Power Basis}
Using the standard monomial basis $\phi = \{1, t, t^2, t^3\}$, where $u(t) = a_0 + a_1t + a_2t^2 + a_3t^3$, the system $Ax=b$ is constructed as:
\begin{equation}
\begin{bmatrix} 
1 & 0 & 0 & 0 \\
1 & 1 & 1 & 1 \\
0 & 0 & 2 & 2 \\
0 & 0 & 2 & 4
\end{bmatrix}
\begin{bmatrix} 
a_0 \\
a_1 \\
a_2 \\
a_3
\end{bmatrix} = 
\begin{bmatrix} 
0 \\
1 \\
0 \\
0
\end{bmatrix}
\end{equation}

The determinant of the matrix $A$ is $4$, indicating a unique solution exists.

The solution to which is:
\begin{equation}
\begin{bmatrix} 
a_0 \\ a_1 \\ a_2 \\ a_3
\end{bmatrix} = 
\begin{bmatrix} 
0 \\ 1 \\ 0 \\ 0
\end{bmatrix}
\end{equation}

The matrix is decomposable into a linear combination of pauli strings.
In this case we need 16 out of 16 total pauli strings to represent the matrix exactly.
% does interchancing rows or colimns change the number of pauli strings needed? Seems like not.


\subsection*{3. Case B: Generalized Basis with Coefficients $C$}
We now define the basis functions $\phi_i(t)$ as general polynomials with coefficients $C_{ij}$:
\begin{align*}
    \phi_0(t) &= C_{00} \\
    \phi_1(t) &= C_{11}t + C_{10} \\
    \phi_2(t) &= C_{22}t^2 + C_{21}t + C_{20} \\
    \phi_3(t) &= C_{33}t^3 + C_{32}t^2 + C_{31}t + C_{30}
\end{align*}
Applying the four conditions to the trial function $u(t) = \sum_{i=0}^3 a_i \phi_i(t)$ yields the system:
\begin{equation}
\begin{bmatrix} 
C_{00} & C_{10} & C_{20} & C_{30} \\ 
C_{00} & (C_{11} + C_{10}) & \sum C_{2j} & \sum C_{3j} \\ 
0 & 0 & 2C_{22} & (2C_{33} + 2C_{32}) \\ 
0 & 0 & 2C_{22} & (4C_{33} + 2C_{32}) 
\end{bmatrix}
\begin{bmatrix} 
a_0 \\ 
a_1 \\ 
a_2 \\ 
a_3 
\end{bmatrix} 
= 
\begin{bmatrix} 
0 \\ 
1 \\ 
0 \\ 
0 
\end{bmatrix}
\end{equation}

The determinant of this matrix is given by:
\begin{equation}
4 C_{00} C_{11} C_{22} C_{33}
\end{equation}

% in general, is the determinant always a product of the leading coefficients?

\subsection*{4. Minimizing the Number of Pauli Strings}

The Pauli decomposition of any $4 \times 4$ matrix $A$ is unique: the coefficient of each
two-qubit Pauli string $P_k$ is determined by
\begin{equation}
    c_k = \frac{\mathrm{Tr}(P_k A)}{4}.
\end{equation}
There are 16 such strings in total.  The goal is to choose the basis coefficients $C_{ij}$
so that as many $c_k$ as possible are exactly zero.  Three structural conditions, imposed
in sequence, achieve this.

\paragraph{Step 1: Enforce block-diagonal structure.}
The upper-right $2\times 2$ block of the general Case B matrix has entries
\begin{equation}
    \begin{pmatrix} C_{20} & C_{30} \\ C_{20}+C_{21}+C_{22} & C_{30}+C_{31}+C_{32}+C_{33} \end{pmatrix}.
\end{equation}
This block is nonzero for generic choices of $C_{ij}$ (e.g.\ $C_{30}=1$ immediately
places a nonzero entry at position $(0,3)$).  To force it to zero we impose:
\begin{equation}
    C_{20} = 0, \qquad C_{30} = 0, \qquad C_{21} + C_{22} = 0, \qquad C_{31}+C_{32}+C_{33} = 0.
\end{equation}
The lower-left block is already zero by construction (the collocation rows evaluate
$\phi_0''$ and $\phi_1''$, which vanish at interior points).  With the above constraints,
the matrix takes the block form
\begin{equation}
    A = \begin{pmatrix} B_1 & 0 \\ 0 & B_2 \end{pmatrix}, \qquad
    B_1 = \begin{bmatrix} C_{00} & C_{10} \\ C_{00} & C_{11}+C_{10} \end{bmatrix}, \quad
    B_2 = \begin{bmatrix} 2C_{22} & 2C_{33}+2C_{32} \\ 2C_{22} & 4C_{33}+2C_{32} \end{bmatrix}.
\end{equation}
Any block-diagonal matrix can be written as
$A = I \otimes \tfrac{B_1+B_2}{2} + Z \otimes \tfrac{B_1-B_2}{2}$,
so only Pauli strings with $I$ or $Z$ on the first qubit can appear.
This eliminates eight strings ($XI, XX, XY, XZ, YI, YX, YY, YZ$), leaving at most 8.

\paragraph{Step 2: Equal blocks (eliminates the $Z\otimes\cdot$ strings).}
If $B_1 = B_2$ then the term $Z \otimes \tfrac{B_1-B_2}{2}$ vanishes entirely, removing
$ZI, ZX, ZY, ZZ$.  This leaves at most 4 strings: $II, IX, IY, IZ$.
Equating $B_1 = B_2$ entry by entry gives:
\begin{equation}
    C_{00} = 2C_{22}, \qquad
    C_{10} = 2C_{33}+2C_{32}, \qquad
    C_{11}+C_{10} = 4C_{33}+2C_{32}.
\end{equation}

\paragraph{Step 3: Repeated first-column entry (eliminates $IY$).}
For a $2\times 2$ matrix $B = \bigl[\begin{smallmatrix}a&b\\a&c\end{smallmatrix}\bigr]$,
the coefficient of $Y$ in its single-qubit Pauli decomposition is
$\mathrm{Tr}(Y B)/2 = i(b-a)/2$.
This vanishes if and only if $b = a$, i.e.\ the off-diagonal entry equals the top-left entry.
For $B_1$ this requires $C_{10} = C_{00}$, and since Step~2 already forces $B_1 = B_2$,
the same condition holds for $B_2$ automatically.
After imposing $C_{10} = C_{00}$, only $II$, $IX$, and $IZ$ remain.

\paragraph{Why three is the minimum.}
With the block $B = \bigl[\begin{smallmatrix}a&a\\a&c\end{smallmatrix}\bigr]$, the three
surviving coefficients are
\begin{equation}
    \alpha = \frac{a+c}{2}, \qquad \beta = a, \qquad \gamma = \frac{a-c}{2},
\end{equation}
so that $A = \alpha\, II + \beta\, IX + \gamma\, IZ$ (before normalisation).
Eliminating $IZ$ would require $a = c$, making $\det(B) = a(c-a) = 0$; the system
would have no unique solution.  Hence three Pauli strings is the minimum compatible
with a non-singular collocation matrix.

\subsection*{5. Resulting Decomposition}

Applying all three conditions with the concrete choice
\begin{align*}
    C_{00} &= 2, \\
    C_{10} &= 2, \quad C_{11} = 2, \\
    C_{20} &= 0, \quad C_{21} = -1, \quad C_{22} = 1, \\
    C_{30} &= 0, \quad C_{31} = -1, \quad C_{32} = 0, \quad C_{33} = 1
\end{align*}
gives $a = 2$, $c = 4$, and the matrix
\begin{equation}
A =
\begin{bmatrix}
2 & 2 & 0 & 0 \\
2 & 4 & 0 & 0 \\
0 & 0 & 2 & 2 \\
0 & 0 & 2 & 4
\end{bmatrix}.
\end{equation}
After normalising by $\|A\| = \sqrt{6a^2+2c^2} = 2\sqrt{14}$, the Pauli decomposition is
\begin{equation}
    \frac{A}{\|A\|} = \frac{3}{2\sqrt{14}}\,II + \frac{1}{\sqrt{14}}\,IX - \frac{1}{2\sqrt{14}}\,IZ
    \;\approx\; 0.4009\,II + 0.2673\,IX - 0.1336\,IZ,
\end{equation}
which is an 81\% reduction from the 16 Pauli strings required in Case A.

\paragraph{Why three is not the global minimum.}
The argument above that three Pauli strings is the minimum rests on the monic
normalisation $C_{jj}=1$ and on the specific block-diagonal structure imposed in
Steps~1--3.  It is the minimum \emph{within that family}, but not over all
admissible collocation matrices.

\subsection*{6. Global Optimum via the Basis-Change Identity}

\paragraph{The identity.}
Any polynomial basis $\{\phi_j\}$ with coefficient matrix $B$
(so that $\phi_j(t)=\sum_{k=0}^{j}B_{jk}\,t^k$) yields a collocation matrix
\begin{equation}
    M = A_{\mathrm{mono}} \cdot B^{\top},
\end{equation}
where $A_{\mathrm{mono}}$ is the Case~A monomial matrix.  Given any invertible
target $M$ we can recover a valid basis via
\begin{equation}
    B = \bigl(A_{\mathrm{mono}}^{-1}\,M\bigr)^{\!\top}.
\end{equation}
Because $B$ is invertible whenever $M$ is, there is a bijection between
non-singular collocation matrices and non-degenerate polynomial bases.
The Pauli count is therefore a property of $M$ alone.

\paragraph{LP formulation.}
Each Pauli coefficient $c_P = \tfrac{1}{4}\operatorname{Tr}(P\,M)$ is linear in the
entries of $M$.  The L1 surrogate for the Pauli count is
\begin{equation}
    \min_{M}\;\sum_{P}|c_P(M)|
    \quad\text{s.t.}\quad \operatorname{Tr}(M) = 4,
\end{equation}
where $\operatorname{Tr}(M)=4$ pins the $II$ coefficient to $c_{II}=1$ and
prevents the trivial solution $M\to 0$.  All 16 matrix entries are free real
variables; the absolute values are linearised with slack variables $t_P\geq|c_P|$,
giving a standard LP solved with the HiGHS simplex method.

\paragraph{Result.}
The LP finds the global optimum
\begin{equation}
    M^* = I_4,
    \qquad
    c_{II} = 1, \quad c_P = 0 \;\text{for all other } P.
\end{equation}
This is a \textbf{single Pauli string}---a 94\% reduction from Case~A and a
further improvement over the three-string analytical solution.

\paragraph{Corresponding basis.}
Applying $B = (A_{\mathrm{mono}}^{-1}\,I_4)^{\top}$ gives the dual basis
\begin{align*}
    \phi_0(t) &= 1 - t, \\
    \phi_1(t) &= t, \\
    \phi_2(t) &= -\tfrac{1}{2}t + t^2 - \tfrac{1}{2}t^3, \\
    \phi_3(t) &= -\tfrac{1}{2}t^2 + \tfrac{1}{2}t^3.
\end{align*}
The basis is non-degenerate ($\det B = \tfrac{1}{4}\neq 0$).  Solving
$M^*\mathbf{a}=\mathbf{b}$ with $\mathbf{b}=[0,1,0,0]^{\top}$ gives
$\mathbf{a}=[0,1,0,0]^{\top}$, so $u(t)=\phi_1(t)=t$, which is the exact solution.

\paragraph{Summary of Pauli counts.}
\begin{center}
\begin{tabular}{lcc}
\hline
Basis & Pauli strings & Notes \\
\hline
Case A (monomial)              & 16 & no structure exploited \\
Case B (block-diagonal, monic) &  3 & analytical construction \\
LP global optimum ($M^*=I_4$)  &  1 & basis-change identity \\
\hline
\end{tabular}
\end{center}

\end{document}


% explain how you got the pauli strings
% run some code numerically justify. Run for two cases. The case B should look better. Add some figures

